$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in the Presence of Non-Functional Properties

### 2014-02-12 JournalToSubmit2013/metamodels.tex

>>>>

Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMs); next,  starting from such models build platform independent models (PIMs) specifying the system details; next transform such models into platform specific models (PSMs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into code that implements the system.

\subsubsection{Computation Independent Models}
$\Pi$-SODM uses two models at the CIM level: The E3value~\cite{e3value} model and BPMN~\cite{BPMN}.
The former  identifies the transference of value information between components of the system.
The BPMN model establishes which are the actors and main tasks of the application.

The e3value defines \textit{dependency paths}, showing the value exchanges, which are triggered by the occurrence of an end-consumer need (in our case, the need of a risk assessment). 
A dependency path has a direction and consists of a sequence of linked dependency nodes.
A dependency path starts with a \textit{start stimulus} node and ends with an \textit{end stimulus} node\footnote{See Legend on Figure~\ref{fig:E3valuemodel}}. 
Dependency paths may also contain \textsl{OR} and \textsl{AND} elements (both for initiate and join alternative and parallel paths).

\subsubsection{Platform Independent Models}
 The \textit{$\pi$-UseCase} meta-model extends the UML Use Case meta-model for describing functional and non-functional requirements. Non-functional requirements are defined as \textit{constraints} over actions and data.

The \textit{$\pi$-ServiceProcess} meta-model defines the concept of \textit{service contract} to represent constraints over data and actions 
%that must be performed upon certain conditions.
%The \textit{$\pi$-ServiceProcess} meta-model gathers the constraints
described in the \textit{$\pi$-UseCase}.
%model into contracts that are associated
%with services.

The \textit{$\pi$-ServiceComposition} meta-model provides the concept  \textit{Policy}
to integrate contracts with similar non-functional requirements.
For instance, security and privacy restrictions may be grouped into a security policy.
%\textit{$\pi$-ServiceComposition} models can be refined into PSMs.

\subsubsection{Platform Specific Models}

The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions. 
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.

===================================

### 2014-02-12 JournalToSubmit2013/relworks.tex

A non-functional property (NFP) specifies criteria about the behavior of a system. These criteria are related to the conditions in which they are executed and to its performance. Non-functional properties are also referred as “constraints”, “quality attributes”, “quality goals”, “quality of service requirements” and “non-behavioral requirements” \cite{StGr05}. In the case of service-based applications, non-functional requirements concern the application itself as well as its component services. 
The majority of the work on NFPs uses a product-oriented approach, which is concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.

@@

\subsection{Classifying Non-functional properties}
We conducted a systematic review \cite{placidoPhDThesis2012} in order to analyze approaches that consider non-functional properties for developing service-oriented applications. As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems. The classification is organized in three layers representing: application modeling, services composition and services. The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP). At the application layer NFP can refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an “@” and a “.”. A value NFR expresses constraints about the way data and functions can be ac-cessed and executed. 
Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).  Value constraints concern more on the conditions in which services can be used. For example, accessing to a function within an authentication protocol.
Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.  For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.

\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. Therefore they propose:
\begin{itemize}
\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
\end{itemize}

NFP concepts are close to those defined by \cite{Souz12} but they are not organized into layers.  They correspond essentially to the service and service composition layers. The enforcement requirements describe the way NFP are weaved to service coordinations. Existing works respect all or a subset of these requirements. The chosen requirements have an impact in the way NFP are enforced at execution time.

@@

Extensions to UML are proposed to express NFPs. NFPs can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}. The performance engineering [DiSD02] annotates quantitative performance constraints to UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram. The UML Profile for Modeling Quality of Service and Fault Tolerance Characteris-tics and Mechanisms \cite{Omg04} describes a set of QoS characteristics such as throughput, latency, integrity, dependability, among others. The characteristics can also be parameter-ized with for example the type of keys or types of encryption algorithms in case of the secu-rity characteristics. For all UML profiles based approaches it is generally possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g. Class or Opera-tion. The ordering of stereotypes cannot be expressed but composition- related require-ments can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams.

\cite{OrHe07} incorporate extra-functional properties into UML models by the use of UML stereotypes. For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
Design patterns help in documenting and communicating proven design solutions to recurring problems. Design patterns not only describe how to solve design problems but also document the designers’ intentions - why the solution is chosen over others and what tradeoffs are made.

\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.

Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application. According to the Busi-ness Rules Group \cite{Hay00}, a business rule is a statement that defines or constrains some aspect of the business. It is intended to assert business structure or to control the behavior of the business \cite{Hay00}. Business rules are usually expressed either as constraints or in the form if conditions then action. The conditions are also called rule premises. The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}. Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making. A business rule system is a system in which the rules are separated logically and perhaps physically from the other parts \cite{ChMe04}. \cite{ChMe04} investigate a hybrid approach, which combines business processes as known e.g., from BPEL with business rules \cite{Hay00}.

@@

The AO4BPEL \cite{CSHM06} deployment descriptor (DD) is used for specifying non-functional requirements for BPEL processes. First the service (one of logging, security, transactions and reliable messaging) is selected. Then the class (e.g. authentication or con-fidentiality for security) can be specified. The type (e.g. encryption or decryption for confi-dentiality) is then the concrete action that must be executed in order to enforce the re-quirement. Then a selector can be defined such as for a certain BPEL activity. Each selector can be mapped to a requirement.

The W3C proposes a set of languages and protocols for specifying the NFP of a ser-vices’ coordination. A classification of NFP protocols, provided by the WS-* family protocols proposed by the World Wide Web Consortium (W3C). 
%These protocols suppose an underlying communication protocol. 
%
%The following sections describe representative examples of the WS-* protocol-based approaches.
%
 Standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols. WSDL and SOAP are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer. As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
 % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
 Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services. In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.

As WS-* and similar approaches, our work enables the specification and programming of crosscutting aspects (i.e., atomicity, security, exception handling, persistence).
In contrast to these approaches, our work specifies policies for a services' composition in an orthogonal way. Besides, these approaches suppose that non-functional properties are implemented according a the knowledge that a programmer has of a specific application requirements but they are not derived in a methodological way, leading to ad-hoc solutions that can be difficult to reuse. In our approach, once defined the policies for a given application they can be reused and/or specialized for another one with the same requirements or that uses services that impose the same constraints.

@@

Several approaches adding non-functional properties to service compositions extend the underlying composition model. Non-functional properties can concern for example securty, atomicity, availability, response time.
\cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
non-functional propoerties are  are modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time. Quality requirements  are then associated to the components of an architecture that implements a service composition.

Other approaches specify architectures or frameworks to manage and validate QoS attributes in service composition processes. 
\cite{XiaoCZBOLH08} proposes a framework to verify NFP compliance in composed services. Tasks in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time. 
\cite{SchmelingCM11} uses a formal method  to choose best suited services instances according to a service composition pattern that includes some QoS attributes. 
 \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services. By executing an algorithm, the framework allow to choose the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime. \cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for includin  QoS attributes that can be evaluated at runtime.
 \cite{Karunamurthy2012787} proposes an architecture in which web services are described with non-functional characteristics that are validated in the service composition process.

@@

\subsection{Methodologies for Service-oriented applications development  including non-func\-tion\-al properties}

@@

Unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that allows to model service-based applications starting from a high-level business modelling, and defining meta-models at different abstraction levels (CIM, PIM and PSM), it enables the design and development of services-based applications that can be reused and that are reliable. The definition of a model-driven approach that allows to model non-functional properties in the early stages of the software development process, and the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefits of our proposal.

>>>>

A non-functional property (NFP) specifies criteria about the behavior of a system. These criteria are related to the conditions in which they are executed and to its performance. Non-functional properties are also referred as “constraints”, “quality attributes”, “quality goals”, “quality of service requirements” and “non-behavioral requirements” \cite{StGr05}. In the case of service-based applications, non-functional requirements concern the application itself as well as its component services. 
The majority of the work on NFPs uses a product-oriented approach concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.

@@

\subsection{Classifying Non-functional properties}
We conducted a systematic review \cite{placidoPhDThesis2012} in order to analyze approaches that consider non-functional properties for developing service-oriented applications. As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems. The classification is organized in three layers representing: application modeling, services composition and services. The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP). At the application layer NFP can refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an “@” and a “.”. A value NFR expresses constraints about the way data and functions can be accessed and executed. 
Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).  Value constraints concern more on the conditions in which services can be used. For example, accessing to a function within an authentication protocol.
Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.  For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.

\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. 
%
%Therefore they propose:
%\begin{itemize}
%\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
%\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
%\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
%\end{itemize}
%
NFP concepts are close to those defined by \cite{Souz12} but they are not organized into layers.  They correspond essentially to the service and service composition layers. The enforcement requirements describe the way NFP are weaved to service coordinations. Existing works respect all or a subset of these requirements. The chosen requirements have an impact in the way NFP are enforced at execution time.

@@

Extensions to UML are proposed to express NFPs that can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}.    UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram are annotated with quantitative performance constraints. The UML Profile for Modeling Quality of Service and Fault Tolerance Characteristics and Mechanisms \cite{Omg04} describe a set of QoS characteristics such as throughput, latency, integrity, and dependability. The characteristics can also be parameterized with for example the type of keys or types of encryption algorithms in case of the security characteristics. For all UML profiles based approaches it is generally possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g. Class or Operation. For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services. The ordering of stereotypes cannot be expressed but composition-related requirements can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams \cite{OrHe07}.

 %incorporate extra-functional properties into UML models by the use of UML stereotypes. 
%Design patterns help in documenting and communicating proven design solutions to recurring problems. Design patterns not only describe how to solve design problems but also document the designers’ intentions - why the solution is chosen over others and what tradeoffs are made.

%\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.

%Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application. According to the Business Rules Group \cite{Hay00}, a business rule  is intended to assert business structure or to control the behavior of the business \cite{Hay00}. Business rules are usually expressed either as constraints or in the form if conditions then action. The conditions are also called rule premises. The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}. Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making. 
In a business rule system  rules are logically separated  and perhaps physically from the other parts \cite{ChMe04}. \cite{ChMe04} investigate a hybrid approach, which combines business processes as known e.g., from BPEL with business rules \cite{Hay00}.

@@

The AO4BPEL \cite{CSHM06} deployment descriptor (DD) is used for specifying non-functional requirements for BPEL processes. First the service (one of logging, security, transactions and reliable messaging) is selected. Then the class (e.g. authentication or confidentiality for security) can be specified. The type (e.g. encryption or decryption for confidentiality) is then the concrete action that must be executed in order to enforce the requirement.  A selector can be defined such as for a certain BPEL activity. Each selector can be mapped to a requirement.

The W3C proposes a set of languages and protocols for specifying the NFP of a ser-vices’ coordination. A classification of NFP protocols, provided by the WS-* family protocols proposed by the World Wide Web Consortium (W3C). 
%These protocols suppose an underlying communication protocol. 
%
%The following sections describe representative examples of the WS-* protocol-based approaches.
%
 Standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols. WSDL and SOAP are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific non-functional constraints to a services composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer.

% As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
% % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
% Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services. In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.

@@

Several approaches adding non-functional properties to service compositions extend the underlying composition model. Non-functional properties can concern for example securty, atomicity, availability, response time.
\cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
non-functional propoerties are   modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time. Quality requirements  are then associated to the components of an architecture that implements a service composition.

Other approaches specify architectures or frameworks to manage and validate QoS attributes in service composition processes. 
\cite{XiaoCZBOLH08} proposes a framework to verify NFP compliance in composed services. Tasks in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time. 
\cite{SchmelingCM11} uses a formal method  to choose best suited services instances according to a service composition pattern that includes some QoS attributes. 
 \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services. By executing an algorithm, the framework  chooses the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime. \cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for including  QoS attributes that can be evaluated at runtime.
 \cite{Karunamurthy2012787} proposes an architecture in which web services are described with non-functional characteristics that are validated in the service composition process.

@@

\subsection{Towards methodologies for developing Service-oriented applications   with non-func\-tion\-al properties}

@@

Unlike methodologies and approaches providing best practices presented above, the main contribution of our proposal is that, integrated to a method that allows to model service-based applications starting from a high-level business modelling, and defining meta-models at different abstraction levels (CIM, PIM and PSM), it enables the design and development of services-based applications that can be reused and that are reliable. The definition of a model-driven approach  allows to model non-functional properties in the early stages of the software development process, and the definition of model transformations to go down in the development process adding more a more implementation details at each modelling stage is a key benefits of our proposal.
